---
title: "Riborex Manual"
author: Wenzheng Li, Weili Wang, Philip J. Uren, Luiz OF Penalva, Andrew D. Smith 
output: pdf_document
---

# Introduction
Riborex is a R package for identifying differentially translated genes from Ribo-seq
data. Riborex integrates both RNA- and Ribo-seq read count data into a single
generalized linear model (GLM) and generates a modified design matrix reflecting
the integration. At its core, Riborex applies existing RNA-seq analysis
tools such as edgeR, DESeq2 and Voom to this modified design matrix and identifies
differential translation across conditions. 

# Detailed example
First, we need to load Riborex library.
```{r message=FALSE}
library(riborex)
```

The input for Riborex are two read count tables summarized from RNA-seq and
Ribo-seq data respectively. The read count table should be organized as a
data frame with rows correspond to genes and columns correspond to samples as
shown below.
```{r}
data(riborexdata)
RNACntTable <- riborexdata$rna
RiboCntTable <- riborexdata$ribo
```

We can check the first five lines of the table:
```{r}
head(RNACntTable,5)
head(RiboCntTable,5)
```

Then we need to prepare two vectors to indicate the treatments of samples in RNA- and Ribo-seq data. 
Both RNA-seq and Ribo-seq can have different number of samples in control
and treated condtions, and RNA-seq and Ribo-seq data can have different number of samples.
```{r}
rnaCond <- c("control", "control", "treated", "treated")
riboCond <- c("control", "control", "treated", "treated")
```

After the two read count table and two condition vectors are ready,
we can use riborex (), and we can choose which engine to use.
By default, DESeq2 is used as the engine if you don't specify the
engine option. Use help(riborex) in R to see more details about this
function.
```{r message=FALSE, warning=FALSE}
res.deseq2 <- riborex(RNACntTable, RiboCntTable, rnaCond, riboCond)
```

The format of the result is the same when DESeq2 is used in RNA-seq analysis.
```{r results='hold'}
res.deseq2
```

Also, you can use summary () for your results.
```{r}
summary(res.deseq2)
```

And results can be saved by:
```{r}
write.table(res.deseq2, "riborex_res_deseq2.txt", quote=FALSE)
```



If you want to use edgeR as your engine, you can use riborex () as:
```{r message=FALSE}
res.edgeR <- riborex(RNACntTable, RiboCntTable, rnaCond, riboCond, "edgeR")
```

The format of the result is the same when edgeR is used in RNA-seq analysis.
```{r}
head(res.edgeR$table)
```

For edgeR engine, you can also choose to estimate dispersion of RNA-seq and Ribo-seq
data separately by specifying engine as "edgeRD".
```{r message=FALSE}
res.edgeRD <- riborex(RNACntTable, RiboCntTable, rnaCond, riboCond, "edgeRD")
```

If you want to use Voom as the engine, you can run riborex () as:
```{r message=FALSE}
res.voom <- riborex(RNACntTable, RiboCntTable, rnaCond, riboCond, "Voom")
```

The format of the result is the same when Voom is used in RNA-seq analysis.
```{r}
head(res.voom)
```